---
title: "TMB: truncal and subclonal; associations with OS"
output: html_notebook
---

Run this notebook to generate
* Figure 1e
* Extended Data Figure 1g-i

This notebook tests associations of pre-ICB TMB with overall survival from ICB start.
- OncoPanel data: assay-reported non-syn TMB
- WES data: truncal non-syn TMB and total non-syn TMB
- Combined OncoPanel and WES: using total non-syn TMB from WES
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
library(survival)
library(survminer)
```

```{r}
wes_tmb_sheet <- read_delim('../data/icb_wes_mut_counts.tsv',
                     show_col_types = FALSE) %>% 
  mutate(clonal_status = gsub('_pair', '', clonal_status))

oncopanel_tmb_sheet <- read_delim('../data/oncopanel_sample_sheet.tsv',
                                  show_col_types = FALSE)

oncopanel_sample_dict <- read_excel('../data/oncopanel_clinical.xlsx') %>% 
  filter(!is.na(id_broad_sample)) %>% 
  select(id_uniquesample, id_broad_sample)

broad_participant_sheet <- read_delim('../data/participant_cohort_sheet.tsv',
                            show_col_types = FALSE)

broad_sample_sheet <- read_delim('../data/sample_cohort_sheet.tsv',
                            show_col_types = FALSE) %>% 
    filter(str_equal(tumor_normal, 'tumor'))
```

Create tidy dataframes
```{r}
wes_samples <- broad_sample_sheet %>%
  filter(usable_wes)

oncopanel_samples <- broad_sample_sheet %>% 
  filter(usable_oncopanel)

oncopanel_tmb_sheet <- oncopanel_tmb_sheet %>% 
  mutate(id_uniquesample = as.character(id_uniquesample)) %>% 
  left_join(oncopanel_sample_dict, by = 'id_uniquesample') %>% 
  mutate(sample_id = ifelse(is.na(id_broad_sample),
                           id_uniquesample, id_broad_sample)) %>% 
  select(sample_id, MUTATIONAL_BURDEN) %>% 
  mutate(sample_id = ifelse(str_equal(sample_id, 'GBM.ICB-3-14.637.Post.FFbank'), 'GBM.ICB-3-14.233.Pre', sample_id))
```

# OncoPanel data: assay-reported TMB
```{r}
oncopanel_latest_pre_tmb <- oncopanel_samples %>% 
  group_by(participant_id_legacy) %>% 
  filter(collection_date_dfd == max(collection_date_dfd)) %>% 
  ungroup %>% 
  left_join(oncopanel_tmb_sheet, by = c('sample_id_legacy' = 'sample_id')) %>% 
  left_join(broad_participant_sheet, by = 'participant_id_legacy') %>% 
  filter(!is.na(MUTATIONAL_BURDEN)) %>% 
  group_by(`ICB for Newly diagnosed`) %>%
  mutate(deceased_bool = str_equal(Deceased, 'Deceased'),
         newly_diagnosed = str_equal(`ICB for Newly diagnosed`, 'Newly-diagnosed'),
         TMB_bin =  factor(ifelse(MUTATIONAL_BURDEN >= median(MUTATIONAL_BURDEN), 'TMB above median', 'TMB below median'), levels = c('TMB above median', 'TMB below median'))) %>%
  ungroup %>%
  select(participant_id_legacy, osicb, deceased_bool, newly_diagnosed, TMB_bin,
         MUTATIONAL_BURDEN)

hist(oncopanel_latest_pre_tmb$MUTATIONAL_BURDEN, breaks = 20)
```

```{r}
fit_oncopanel_ND <- survfit(Surv(osicb, deceased_bool) ~ TMB_bin,
               data = oncopanel_latest_pre_tmb %>% filter(newly_diagnosed))

fit_oncopanel_R <- survfit(Surv(osicb, deceased_bool) ~ TMB_bin,
               data = oncopanel_latest_pre_tmb %>% filter(!newly_diagnosed))

p_ND <- ggsurvplot(fit_oncopanel_ND,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           ggtheme = theme_classic(),
           palette = c('#7A6288', '#C7AFD5'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = paste0('Newly diagnosed OncoPanel cohort\n(n = ',
                          oncopanel_latest_pre_tmb %>% filter(newly_diagnosed) %>% dim %>% .[1],
                          ')')
           ) %++% guides(color = guide_legend(nrow = 2))

p_R <- ggsurvplot(fit_oncopanel_R,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           ggtheme = theme_classic(),
           palette = c('#588556', '#A5D2A2'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = paste0('Recurrent OncoPanel cohort\n(n = ',
                          oncopanel_latest_pre_tmb %>% filter(!newly_diagnosed) %>% dim %>% .[1],
                          ')')
           ) %++% guides(color = guide_legend(nrow = 2))

p_combined <- ggarrange(p_ND$plot, p_R$plot,
                        common.legend = FALSE)

p_combined

```

# WES data
```{r}
pre_wes_pts <- wes_samples %>%
  filter(str_equal(pre_post, 'Pre')) %>% 
  pull(participant_id_legacy) %>% 
  unique

wes_latest_pre_samples <- wes_samples %>% 
  filter(str_equal(pre_post, 'Pre')) %>% 
  group_by(participant_id_legacy) %>% 
  filter(collection_date_dfd == max(collection_date_dfd)) %>%
  pull(sample_id_legacy)

wes_truncal_tmb <- wes_tmb_sheet %>%
  filter(Patient_ID %in% pre_wes_pts & str_equal(clonal_status, 'clonal')) %>% 
  select(Patient_ID, tmb) %>% 
  rename(truncal_TMB = tmb)

wes_latest_pre_tmb <- wes_tmb_sheet %>% 
  filter(clonal_status %in% wes_latest_pre_samples) %>% 
  group_by(Patient_ID) %>% 
  summarize(pre_TMB = mean(tmb))

wes_tmb <- wes_truncal_tmb %>%
  left_join(wes_latest_pre_tmb, by = 'Patient_ID')

wes_tmb <- wes_tmb %>%
  left_join(broad_participant_sheet, by = c('Patient_ID' = 'participant_id_legacy')) %>%
  group_by(`ICB for Newly diagnosed`) %>%
  mutate(deceased_bool = str_equal(Deceased, 'Deceased'),
         newly_diagnosed = str_equal(`ICB for Newly diagnosed`, 'Newly-diagnosed'),
         truncal_TMB_bin = factor(ifelse(truncal_TMB >= median(truncal_TMB), 'TMB above median', 'TMB below median'), levels = c('TMB above median', 'TMB below median')),
         pre_TMB_bin = factor(ifelse(pre_TMB >= median(pre_TMB), 'TMB above median', 'TMB below median'), levels = c('TMB above median', 'TMB below median'))) %>%
  ungroup %>%
  select(Patient_ID, truncal_TMB, truncal_TMB_bin, pre_TMB, pre_TMB_bin, osicb, deceased_bool, newly_diagnosed)

n_distinct(wes_tmb$Patient_ID)
```

```{r}

fit_wes_pre_ND <- survfit(Surv(osicb, deceased_bool) ~ pre_TMB_bin,
               data = wes_tmb %>% filter(newly_diagnosed))

fit_wes_pre_R <- survfit(Surv(osicb, deceased_bool) ~ pre_TMB_bin,
               data = wes_tmb %>% filter(!newly_diagnosed))

p_pre_ND <- ggsurvplot(fit_wes_pre_ND,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           ggtheme = theme_classic(),
           palette = c('#7A6288', '#C7AFD5'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = paste0('Newly diagnosed exome cohort\n(n = ',
                          wes_tmb %>% filter(newly_diagnosed) %>% dim %>% .[1],
                          ')')
           ) %++%
  guides(color = guide_legend(nrow = 2))

p_pre_R <- ggsurvplot(fit_wes_pre_R,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           ggtheme = theme_classic(),
           palette = c('#588556', '#A5D2A2'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = paste0('Recurrent exome cohort\n(n = ',
                          wes_tmb %>% filter(!newly_diagnosed) %>% dim %>% .[1],
                          ')')
           ) %++%
  guides(color = guide_legend(nrow = 2))

p_combined <- ggarrange(p_pre_ND$plot, p_pre_R$plot,
                        common.legend = FALSE)

p_combined

```

# Combining WES and OncoPanel
Pre-ICB overall TMB only; we will not stratify by truncal/subclonal status for
this comparison.
```{r}
combined_tmb <- oncopanel_latest_pre_tmb %>% 
  full_join(wes_tmb, by = c('participant_id_legacy' = 'Patient_ID',
                            'deceased_bool' = 'deceased_bool',
                            'newly_diagnosed' = 'newly_diagnosed',
                            'osicb' = 'osicb')) %>%
  group_by(newly_diagnosed) %>% 
  mutate(TMB_harmonized = ifelse(is.na(MUTATIONAL_BURDEN), pre_TMB, MUTATIONAL_BURDEN),
         median_TMB = median(TMB_harmonized),
         TMB_harmonized_bin = factor(ifelse(TMB_harmonized >= median_TMB, 'TMB above median', 'TMB below median'), levels = c('TMB above median', 'TMB below median'))) %>% 
  ungroup

#combined_tmb %>% write_delim('../../../results/2024-06-26_tmb/combined_tmb.tsv',
#                             delim = '\t')

p_compare_1 <- ggscatter(combined_tmb %>% drop_na,
          x = 'MUTATIONAL_BURDEN',
          y = 'pre_TMB',
          add = 'reg.line',
          xlab = 'OncoPanel TMB',
          ylab = 'Exome TMB',
          conf.int = TRUE,
          title = paste0('All patients with pre-ICB\nOncoPanel and WES\n(n = ',
                         combined_tmb %>% drop_na %>% dim %>% .[1],
                         ')')) +
  stat_cor(method = 'spearman', cor.coef.name = 'rho') +
  geom_abline(aes(slope = 1, intercept = 0, color = 'y = x')) +
  scale_color_manual(values = 'red') +
  labs(colour = '') +
  coord_fixed()

p_compare_2 <- ggscatter(combined_tmb %>% drop_na %>% filter(pre_TMB < 60),
          x = 'MUTATIONAL_BURDEN',
          y = 'pre_TMB',
          add = 'reg.line',
          xlab = 'OncoPanel TMB',
          ylab = 'Exome TMB',
          conf.int = TRUE,
          title = paste0('Patients with\npre-ICB exome TMB < 60 mut/Mb\n(n = ',
                         combined_tmb %>% drop_na %>% filter(pre_TMB < 60) %>% dim %>% .[1],
                         ')')) +
  stat_cor(method = 'spearman', cor.coef.name = 'rho') +
  geom_abline(aes(slope = 1, intercept = 0, color = 'y = x')) +
  scale_color_manual(values = 'red') +
  labs(colour = '') +
  coord_fixed()

p_compare <- ggarrange(p_compare_1, p_compare_2,
                       common.legend = TRUE,
                       legend = 'bottom')
p_compare

```

```{r}
fit_harmonized_ND <- survfit(Surv(osicb, deceased_bool) ~ TMB_harmonized_bin,
               data = combined_tmb %>% filter(newly_diagnosed))
fit_harmonized_R <- survfit(Surv(osicb, deceased_bool) ~ TMB_harmonized_bin,
               data = combined_tmb %>% filter(!newly_diagnosed))

ND_title <- paste0('Newly diagnosed cohort\nmedian TMB = ',
                   combined_tmb %>% filter(newly_diagnosed) %>% pull(median_TMB) %>% unique %>% round(., 1),
                   ' mut/Mb')

R_title <- paste0('Recurrent cohort\nmedian TMB = ',
                  combined_tmb %>% filter(!newly_diagnosed) %>% pull(median_TMB) %>% unique %>% round(., 1),
                  ' mut/Mb')

p_harmonized_ND <- ggsurvplot(fit_harmonized_ND,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           pval.coord = c(1000, 0.6),
           ggtheme = theme_classic(),
           palette = c('#7A6288', '#C7AFD5'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = ND_title
           ) %++%
  guides(color = guide_legend(nrow = 2))

p_harmonized_R <- ggsurvplot(fit_harmonized_R,
           pval = TRUE,
           conf.int = TRUE,
           xscale = 'd_y',
           break.x.by = 365.25,
           xlab = 'Time from ICB start (years)',
           ylab = 'Fraction surviving',
           pval.coord = c(1000, 0.6),
           ggtheme = theme_classic(),
           palette = c('#588556', '#A5D2A2'),
           legend = 'bottom',
           legend.title = '',
           legend.labs = c('TMB above median', 'TMB below median'),
           title = R_title
           ) %++%
  guides(color = guide_legend(nrow = 2))

p_combined <- ggarrange(p_harmonized_ND$plot, p_harmonized_R$plot,
                        common.legend = FALSE)

p_combined

```
