---
title: "Extended_figure_7_snRNAseq_data"
format: html
editor: visual
---

Run this notebook to generate

* Figure 2i, 2j

* Extended Data Figure 7d

* Extended Data Figure 8a, 8b, 8c, 8d, 8e

This script analyzes single-nucleus RNA-seq (snRNA-seq) pseudobulk data to investigate the cellular origins of Verhaak transcriptional subtypes and HLA expression, and their association with survival in the ICB cohort.

- Bivariate Cox survival analysis of MES scores derived from malignant vs. TME compartments.

- Heatmap of Verhaak signature genes correlating snRNA-seq pseudobulk with bulk RNA subtypes.

- Analysis of the fractional contribution of each cell type to the overall MES signature.

- Comparison of HLA class I/II scores and gene expression across malignant cell states.

- Analysis of the cellular sources of HLA class I/II signals, stratified by MES vs. non-MES tumors.


Load libraries, data and plotting themes

```{r setup, include=FALSE}

# load libraries
library(tidyverse)
library(rstatix)
library(ggpubr)
library(ggh4x)
library(scales)
library(dendextend)
library(patchwork)
library(stringr)
library(survival)
library(coxphf)


# load data
metadata_ICB <- readRDS(file = "../data/metadata_ICB.rds")
metadata_ICB %>% dim() # 666948  15

project_metadata <- readRDS(file = "../data/Single cell portal/project_metadata_ICB.rds")

pseudobulk_pre_path <- "../data/pseudobulks_by_sample_cell_type_pre_TPM.rds"
pseudobulks_pre_tpm <- readRDS(pseudobulk_pre_path)


# generate ggplot themes
base_theme <- function() {
  
  theme_classic(base_size = 25) + 
    theme(plot.title = element_text(face = "bold"), plot.title.position = "plot", plot.subtitle = element_text(lineheight = 1.1),
          axis.line = element_blank()
    )
  
}

umap_theme <- function() {
  
  base_theme() +
    theme(axis.ticks = element_blank(), axis.text = element_blank(),
          axis.title.x = element_text(hjust = 0.1), axis.title.y = element_text(hjust = 0.1), axis.title = element_text(size = 14)
    )
  
}

default_theme <- function() {
  
  theme_classic(base_size = 25) + 
    theme(plot.title = element_text(face = "bold"), plot.title.position = "plot", plot.subtitle = element_text(lineheight = 1.1),
          strip.text = element_text(size = 20),
          panel.grid.minor.y = element_line(linewidth = 0.25, linetype = 1),
          panel.grid.major.y = element_line(linewidth = 0.5, linetype = 1),
    )
  
}

default_theme_border <- function() {
  
  default_theme() + 
    theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=1), axis.line = element_blank())
  
}

default_theme_xrot_90 <- function() {
  
  default_theme() +
    theme(axis.text.x = element_text(vjust = 0.5, angle = 90, hjust = 1))
}

default_theme_xrot_90_border <- function() {
  
  default_theme() +
    theme(axis.text.x = element_text(vjust = 0.5, angle = 90, hjust = 1),
          panel.border = element_rect(colour = "black", fill=NA, linewidth=1), axis.line = element_blank())
}


# load color palettes
HEATMAP_LOW_VALUE_C <- "#3C5488"
HEATMAP_HIGH_VALUE_C <- "#E64B35"

SURVIVAL_LS_C <- "#D494B9" 
SURVIVAL_SS_C <- "#32B18F" 

SURVIVAL_LS_MALIGNANT_C <- "#B22222" 
SURVIVAL_SS_MALIGNANT_C <- "#F08080"

SURVIVAL_LS_NON_MALIGNANT_C <- "#2C6BA0" 
SURVIVAL_SS_NON_MALIGNANT_C <- "#9CC9FF" 

RX_UNTREATED_C <- "#4A6274"
RX_TREATED_C <- "#E294AF" 

RX_PRE_ICB_C <- "#4089B2"
RX_POST_ICB_C <- "#DC8848"


VERHAAK_SUBTYPE_MESENCHYMAL_C <- "#D62727"
VERHAAK_SUBTYPE_CLASSICAL_C <- "#F39354"
VERHAAK_SUBTYPE_NEURAL_C <- "#96A6AF"
VERHAAK_SUBTYPE_PRONEURAL_C <- "#1F73AF"


GBM_STATE_AC_C <- "#FFCC80"
GBM_STATE_GPC_C <- "#B39DDB"
GBM_STATE_MES_C <- "#EF9A9A"
GBM_STATE_NPC_C <- "#B0BEC5"
GBM_STATE_OPC_C <- "#90CAF9"
GBM_STATE_ND_C <- "white"

CELL_TYPE_MALIGNANT_C <- "#C2185B"
CELL_TYPE_OLIGODENDROCYTE_C <- "#FF6961"
CELL_TYPE_MYELOID_C <- "#FFB480"
CELL_TYPE_ASTROCYTES_C <- "#F8F38D"
CELL_TYPE_NEURONS_C <- "#D4E157"
CELL_TYPE_ENDOTHELIAL_C <- "#81C784"
CELL_TYPE_PERICYTES_C <- "#42d6a4"
CELL_TYPE_T_CELL_C <- "#59adf6"
CELL_TYPE_NK_CELL_C <- "#9d94ff"


# create utilities functions
`%!in%` <- Negate(`%in%`)

```

# Figure 2i

MES Score Bivariate Survival Analysis with Forest Plot. Calculate MES scores over pseudo-bulk objects (on a per-sample basis), run bivariate Cox model, create forest plot.

```{r Fig.2i-MES-signal-source-vs-survival}

library(coxphf)
library(survival)

VERHAAK_MES_GENES <- c("ACP3","ACSL1","ADAM12","ALDH3B1","ALOX5","AMPD3","ANXA1","ANXA2",
                       "ANXA4","ARHGAP29","ARPC1B","ARSJ","ASL","BATF","BDKRB2","BLVRB","BNC2",
                       "C1orf54","C5AR1","CASP1","CASP4","CASP5","CASP8","CAST","CAVIN1",
                       "CCR5","CD14","CD2AP","CD4","CDCP1","CEBPB","CHI3L1","CLCF1","CLEC2B",
                       "CLIC1","CNN2","COL1A1","COL1A2","COL5A1","COL8A2","COLGALT1","COPZ2",
                       "CPQ","CRYBG1","CSTA","CTSB","CTSC","CTSZ","CYBRD1","CYTH4","CYTIP",
                       "DAB2","DCBLD2","DOK3","DRAM1","DSC2","DSE","EFEMP2","EHD2","ELF4",
                       "EMP3","ENG","FCGR2A","FCGR2B","FES","FHL2","FHOD1","FMNL1","FNDC3B",
                       "FOLR2","FPR3","FURIN","FXYD5","GCNT1","GNA15","GRN","HEXA","HEXB",
                       "HFE","HK3","ICAM3","IFI30","IGFBP6","IL15RA","IL1R1","IL4R","IQGAP1",
                       "ITGA4","ITGA5","ITGAM","ITGB2","KYNU","LAIR1","LAMB1","LAPTM5","LCP1",
                       "LCP2","LGALS1","LGALS3","LHFPL2","LILRB2","LILRB3","LOX","LRRFIP1",
                       "LTBP1","LTBP2","LY75","LY96","MAFB","MAN1A1","MAN2A1","MAN2B1",
                       "MAPK13","MCUB","MFSD1","MGAT1","MGST2","MRC2","MS4A4A","MSR1","MVP",
                       "MYH9","MYO1F","MYOF","NCF2","NCF4","NOD2","NPC2","NRP1","P4HA2","PDPN",
                       "PHF11","PIGP","PLA2G15","PLAU","PLAUR","PLBD1","PLK3","PLS3","POLD4",
                       "PROCR","PTGER4","PTPN22","PTPN6","PTPRC","PYGL","RAB11FIP1","RAB27A",
                       "RAB32","RABGAP1L","RAC2","RBKS","RBM47","RBMS1","RELB","RHOG","RRAS",
                       "RUNX2","S100A11","S100A13","S100A4","SAT1","SCPEP1","SEC24D",
                       "SERPINA1","SERPINE1","SFT2D2","SH2B3","SHC1","SIGLEC7","SIGLEC9",
                       "SLAMF8","SLC10A3","SLC11A1","SLC16A3","SP100","SQOR","SRPX2","ST14",
                       "STAB1","STAT6","STXBP2","SWAP70","SYNGR2","SYPL1","TCIRG1","TEC","TES",
                       "TGFBI","TGFBR2","TGOLN2","THBD","THBS1","THEMIS2","TIMP1","TLR2",
                       "TLR4","TMBIM1","TNFAIP3","TNFAIP8","TNFRSF11A","TNFRSF1A","TNFRSF1B",
                       "TRADD","TRIM22","TRIM38","TRPM2","TYMP","UAP1","UCP2","VAMP5","VDR",
                       "WIPF1","WWTR1","YAP1","ZNF217")



# genereate malignant cells pseudobulks
pseudobulks_malignant_by_sample_pre_TPM <-
  pseudobulks_pre_tpm %>% 
  reshape2::melt(varnames = c("gene_name", "sample_cell_type_id"), value.name = "exp") %>% 
  tidyr::separate(col = "sample_cell_type_id", into = c("Sample_ID", "Cell_type"), sep = "[*]") %>% 
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::select(Cell_type, malignant_group) %>% unique(), by = "Cell_type") %>% 
  dplyr::filter(malignant_group == "Malignant") %>% droplevels() %>% 
  dplyr::group_by(gene_name, Sample_ID) %>% 
  dplyr::summarise(Exp = sum(exp), .groups = "drop") %>% 
  tidyr::pivot_wider(id_cols = c("gene_name"), names_from = "Sample_ID", values_from = "Exp", values_fill = 0) %>% 
  tibble::column_to_rownames(var = "gene_name")

pseudobulks_malignant_by_sample_pre_TPM %>% dim() # 36601  22


# genereate non-malignant cells pseudobulks
pseudobulks_non_malignant_by_sample_pre_TPM <-
  pseudobulks_pre_tpm %>% 
  reshape2::melt(varnames = c("gene_name", "sample_cell_type_id"), value.name = "exp") %>% 
  tidyr::separate(col = "sample_cell_type_id", into = c("Sample_ID", "Cell_type"), sep = "[*]") %>% 
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::select(Cell_type, malignant_group) %>% unique(), by = "Cell_type") %>% 
  dplyr::filter(malignant_group == "Non-malignant") %>% droplevels() %>% 
  dplyr::group_by(gene_name, Sample_ID) %>% 
  dplyr::summarise(Exp = sum(exp), .groups = "drop") %>% 
  tidyr::pivot_wider(id_cols = c("gene_name"), names_from = "Sample_ID", values_from = "Exp", values_fill = 0) %>% 
  tibble::column_to_rownames(var = "gene_name")

pseudobulks_non_malignant_by_sample_pre_TPM %>% dim() # 36601  22


all(colnames(pseudobulks_non_malignant_by_sample_pre_TPM) == colnames(pseudobulks_malignant_by_sample_pre_TPM)) # TRUE
all(rownames(pseudobulks_non_malignant_by_sample_pre_TPM) == rownames(pseudobulks_malignant_by_sample_pre_TPM)) # TRUE


colnames(pseudobulks_malignant_by_sample_pre_TPM) <- paste0(colnames(pseudobulks_malignant_by_sample_pre_TPM), "*Malignant")
colnames(pseudobulks_non_malignant_by_sample_pre_TPM) <- paste0(colnames(pseudobulks_non_malignant_by_sample_pre_TPM), "*Non-malignant")


# score the pseudobulk by the MESENCHYMAL gene set
pseudobulks_malignant_group_by_sample_pre_verhaak_scores <-
  cbind(log2(pseudobulks_malignant_by_sample_pre_TPM + 1),
        log2(pseudobulks_non_malignant_by_sample_pre_TPM + 1)) %>% 
  scalop::sigScores(sigs = list("Mesenchymal" = VERHAAK_MES_GENES), expr.center = FALSE) %>% 
  tibble::rownames_to_column(var = "sample_malignant_group_id") %>% 
  separate(col = "sample_malignant_group_id", into = c("Sample_ID", "malignant_group"), sep = "\\*") %>%
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::select(Patient_ID, Sample_ID) %>% unique(), by = "Sample_ID")


PATIENTS_TO_EXCLUDE <- c("28", "182")

# average scores over patient samples, and exclude patients with over 180 days from ICB treatment to tissue sampling
pseudobulks_malignant_group_by_patient_pre_verhaak_MES_scores <- 
  pseudobulks_malignant_group_by_sample_pre_verhaak_scores %>% 
  dplyr::filter(Patient_ID %!in% PATIENTS_TO_EXCLUDE) %>% 
  dplyr::select(Patient_ID, Sample_ID, malignant_group, Mesenchymal) %>% 
  dplyr::group_by(malignant_group, Patient_ID) %>% 
  dplyr::summarise(mean_MES_score = mean(Mesenchymal), .groups = "drop")


# add patient survival information
pseudobulks_malignant_group_by_patient_pre_verhaak_MES_summary <-
  pseudobulks_malignant_group_by_patient_pre_verhaak_MES_scores %>% 
  tidyr::pivot_wider(id_cols = c("Patient_ID"), names_from = "malignant_group", values_from = "mean_MES_score") %>% 
  dplyr::left_join(project_metadata %>% 
                     dplyr::select(Patient_ID, OS_ICB, Overall_survival_18m), by = "Patient_ID") %>% 
  dplyr::mutate(status = 1)


# calculate bi-variate Firth's COX-regression model
firth_mes_only <- coxphf(Surv(OS_ICB, status) ~ Malignant + `Non-malignant`, 
                         data = pseudobulks_malignant_group_by_patient_pre_verhaak_MES_summary)


bivariate_results <- 
  data.frame(
    term = c("Malignant MES", "TME MES"),
    hr = exp(firth_mes_only$coefficients),
    conf.low = firth_mes_only$ci.lower,
    conf.high = firth_mes_only$ci.upper,
    p.value = firth_mes_only$prob,
    n = rep(firth_mes_only$n, 2)) %>% 
  rstatix::add_significance(p.col = "p.value")


# Define capping limits for visualization
LOWER_CAPPING_LIMIT <- 0.2
UPPER_CAPPING_LIMIT <- 2.0


plot_data <- bivariate_results %>%
  mutate(
    # Create display strings
    hr_ci_str = sprintf("%.2f (%.2f–%.2f)", hr, conf.low, conf.high),
    p_val_str = scales::pvalue(p.value, accuracy = 0.001, add_p = TRUE),
    
    # Add significance indicator
    significance = ifelse(p.value < 0.05, "*", ""),
    
    # Create ordered factor for predictors (reverse order for top-to-bottom)
    term_display = factor(term, levels = c("TME MES", "Malignant MES")),
    
    # Capping logic for confidence intervals
    is_capped_low = conf.low < LOWER_CAPPING_LIMIT,
    conf.low_capped = ifelse(is_capped_low, LOWER_CAPPING_LIMIT, conf.low),
    is_capped_high = conf.high > UPPER_CAPPING_LIMIT,
    conf.high_capped = ifelse(is_capped_high, UPPER_CAPPING_LIMIT, conf.high)
  )



p_table <- ggplot(plot_data, aes(y = term_display)) +
  geom_text(aes(x = 0, label = n), hjust = 0.5, size = 4.5) +
  geom_text(aes(x = 1, label = hr_ci_str), hjust = 0.5, size = 4.5) +
  geom_text(aes(x = 2, label = p_val_str), hjust = 0.5, size = 4.5) +
  annotate("text", x = 0, y = nlevels(plot_data$term_display) + 0.5, 
           label = "N", fontface = "bold", size = 5) +
  annotate("text", x = 1, y = nlevels(plot_data$term_display) + 0.5, 
           label = "HR (95% CI)", fontface = "bold", size = 5) +
  annotate("text", x = 2, y = nlevels(plot_data$term_display) + 0.5, 
           label = "P-value", fontface = "bold", size = 5) +
  theme_void() +
  scale_y_discrete(expand = expansion(add = c(0.5, 1.0))) +
  xlim(-0.5, 2.5)

# Forest plot with error bars and point estimates
p_forest <- ggplot(plot_data, aes(x = hr, y = term_display)) +
  # Reference line at HR = 1
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "grey70", linewidth = 1.0) +
  
  # Confidence interval error bars
  geom_errorbarh(aes(xmin = conf.low_capped, xmax = conf.high_capped), 
                 height = 0.3, linewidth = 1.2) +
  
  # Arrows for capped lower limits
  geom_segment(data = . %>% filter(is_capped_low),
               aes(x = conf.low_capped, xend = conf.low_capped, 
                   y = term_display, yend = term_display),
               arrow = arrow(length = unit(0.2, "cm"), type = "open", ends = "first"),
               lineend = "round", linewidth = 1.2) +
  
  # Arrows for capped upper limits
  geom_segment(data = . %>% filter(is_capped_high),
               aes(x = conf.high_capped, xend = conf.high_capped, 
                   y = term_display, yend = term_display),
               arrow = arrow(length = unit(0.2, "cm"), type = "open", ends = "last"),
               lineend = "round", linewidth = 1.2) +
  
  # Point estimates with red fill
  geom_point(size = 5, shape = 21, fill = "firebrick", color = "black", stroke = 1) +
  
  # Log scale for x-axis
  scale_x_log10(
    breaks = c(0.2, 0.3, 0.5, 0.75, 1.0, 1.5, 2.0),
    labels = c("0.20", "0.30", "0.50", "0.75", "1.00", "1.50", "2.00")
  ) +
  coord_cartesian(xlim = c(0.18, 2.2)) +
  
  # Labels and theme
  labs(
    x = "Hazard Ratio (HR) per 1 SD Increase (log scale)", 
    y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_line(color = "grey92", linetype = "dashed"),
    axis.text.y = element_text(size = 13, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold", margin = margin(t = 10))
  )

# ---------------------------------------------------------------------------
# --- STEP 14: COMBINE AND SAVE FOREST PLOT ---
# ---------------------------------------------------------------------------

final_forest_plot <- p_forest + p_table +
  plot_layout(widths = c(2.5, 2)) +
  plot_annotation(
    title = "Independent Effects of Malignant and TME MES Scores on Survival",
    subtitle = "Bivariate Firth's Cox Regression Model (Both MES Scores, No Covariates)",
    caption = paste(
      sprintf("n = %d patients | Pre-ICB samples | Patients 28 & 182 excluded", unique(plot_data$n)),
      "MES scores calculated using scalop::sigScores (Verhaak Mesenchymal signature)",
      "HR per 1 standard deviation increase in MES score",
      "* p < 0.05 indicates statistical significance",
      sep = "\n"
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5, margin = margin(b = 5)),
      plot.subtitle = element_text(size = 13, hjust = 0.5, margin = margin(b = 10)),
      plot.caption = element_text(hjust = 0, size = 10, color = "grey40", margin = margin(t = 10))
    )
  )

print(final_forest_plot)

```

# Figure 2j

Compare HLA class I scores across different malignant cell-states over pseudo-bulk objects

```{r Fig.2j-HLA-class-I-scores-by-malignant-subtype}

library(Matrix)
library(scalop)

exp_mat_all_tpm_HLA <- readRDS("../data/exp_mat_all_tpm_HLA.rds")

HLA_CLASS_I_GENES <- c("B2M", "HLA-A", "HLA-B", "HLA-C", "HLA-E", "HLA-F", "HLA-G", "HLA-H")
genes_present <- HLA_CLASS_I_GENES[HLA_CLASS_I_GENES %in% rownames(exp_mat_all_tpm_HLA)]

MIN_NUMBER_OF_CELL_PER_CELL_TYPE <- 5

# FILTER TO PRE-ICB SAMPLES IMMEDIATELY
sample_cell_type_counts_summary <- 
  metadata_ICB %>%
  dplyr::filter(Rx == "Pre-ICB") %>%  # <- FILTER UPFRONT HERE
  group_by(Rx, Patient_ID, Sample_ID, malignant_group, Cell_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  dplyr::filter(count >= MIN_NUMBER_OF_CELL_PER_CELL_TYPE)

# Calculate pseudobulk for ALL malignant cell types (including "Malignant (ND)")
# Now only using Pre-ICB samples
bulk_expression_by_cell_type_patient_malignant <- 
  sample_cell_type_counts_summary %>%
  dplyr::filter(malignant_group == "Malignant") %>%
  droplevels() %>%
  pull(Cell_type) %>% 
  levels() %>%
  map(function(cur_cell_type) {
    
    cat("Processing:", cur_cell_type, "\n")
    
    cur_cell_type_counts_summary <- sample_cell_type_counts_summary %>%
      dplyr::filter(Cell_type == cur_cell_type) %>% 
      droplevels()
    
    if (nrow(cur_cell_type_counts_summary) == 0) {
      return(NULL)
    }
    
    bulk_expression_cur_cell_type <- 
      cur_cell_type_counts_summary$Sample_ID %>% unique() %>%
      map(function(cur_cell_type_sample) {
        
        cur_cell_type_sample_cell_id <- metadata_ICB %>%
          dplyr::filter(Cell_type == cur_cell_type, Sample_ID == cur_cell_type_sample) %>%
          pull(Cell_ID)
        
        exp_subset <- exp_mat_all_tpm_HLA[, cur_cell_type_sample_cell_id, drop = FALSE]
        log2(Matrix::rowMeans(exp_subset) + 1)
        
      }) %>% do.call(cbind, .)
    
    colnames(bulk_expression_cur_cell_type) <- cur_cell_type_counts_summary$Sample_ID %>% unique()
    
    bulk_expression_cur_cell_type %>%
      reshape2::melt(varnames = c("gene_name", "Sample_ID"), value.name = "Exp") %>%
      left_join(sample_cell_type_counts_summary %>%
                  dplyr::filter(Cell_type == cur_cell_type) %>% 
                  droplevels() %>%
                  dplyr::select(Rx, Patient_ID, Sample_ID) %>% 
                  unique(), 
                by = "Sample_ID") %>%
      group_by(Rx, Patient_ID, gene_name) %>%
      summarise(mean_exp = mean(Exp), .groups = "drop") %>%
      mutate(Cell_type = cur_cell_type)
    
  }) %>% 
  keep(~ !is.null(.)) %>%
  do.call(rbind, .) %>%
  mutate_at(c("gene_name"), as.character)

# EXCLUDE "Malignant (ND)" BEFORE centering
# Now centering only across Pre-ICB samples
bulk_expression_centered_matrix <- 
  bulk_expression_by_cell_type_patient_malignant %>%
  dplyr::filter(Cell_type != "Malignant (ND)") %>%  # FILTER BEFORE CENTERING
  mutate(rx_patient_cell_type_id = paste0(Rx, "*", Patient_ID, "*", Cell_type)) %>%
  pivot_wider(id_cols = c("gene_name"), 
              names_from = "rx_patient_cell_type_id", 
              values_from = "mean_exp") %>%
  column_to_rownames(var = "gene_name") %>%
  as.matrix() %>%
  scalop::rowcenter()  # CENTERING ONLY PRE-ICB DATA

# Calculate scores
sig_scores <- bulk_expression_centered_matrix %>%
  as.data.frame() %>%
  rownames_to_column("gene_name") %>%
  dplyr::filter(gene_name %in% genes_present) %>%
  pivot_longer(-gene_name, names_to = "id", values_to = "centered_exp") %>%
  group_by(id) %>%
  summarise(HLA_score = mean(centered_exp, na.rm = TRUE), .groups = "drop") %>%
  separate(id, into = c("Rx", "Patient_ID", "Cell_type"), sep = "\\*")
  # NO FILTER HERE - already only Pre-ICB data

# Order cell types by median HLA score (highest to lowest)
cell_type_order <- sig_scores %>%
  group_by(Cell_type) %>%
  summarise(median_score = median(HLA_score, na.rm = TRUE)) %>%
  arrange(desc(median_score)) %>%
  pull(Cell_type)

sig_scores <- sig_scores %>%
  mutate(Cell_type = factor(Cell_type, levels = cell_type_order))

# --- 6. STATISTICS ---
my_comparisons <- list(
  c("MES-like", "AC-like"),
  c("MES-like", "GPC-like"),
  c("MES-like", "NPC-like"),
  c("MES-like", "OPC-like")
)

unpaired_results <- map_dfr(my_comparisons, function(comp) {
  group1 <- comp[1]
  group2 <- comp[2]
  
  data1 <- sig_scores %>% filter(Cell_type == group1) %>% pull(HLA_score)
  data2 <- sig_scores %>% filter(Cell_type == group2) %>% pull(HLA_score)
  
  test <- t.test(data1, data2, paired = FALSE)
  
  data.frame(
    comparison = paste(group1, "vs", group2),
    group1 = group1,
    group2 = group2,
    n1 = length(data1),
    n2 = length(data2),
    p_value = test$p.value
  )
})

paired_results <- map_dfr(my_comparisons, function(comp) {
  group1 <- comp[1]
  group2 <- comp[2]
  
  patients_with_both <- sig_scores %>%
    filter(Cell_type %in% c(group1, group2)) %>%
    group_by(Patient_ID) %>%
    summarise(n_types = n_distinct(Cell_type), .groups = "drop") %>%
    filter(n_types == 2) %>%
    pull(Patient_ID)
  
  if (length(patients_with_both) < 3) {
    return(data.frame(
      comparison = paste(group1, "vs", group2),
      group1 = group1,
      group2 = group2,
      n_paired = length(patients_with_both),
      p_value = NA
    ))
  }
  
  paired_data <- sig_scores %>%
    filter(Patient_ID %in% patients_with_both, Cell_type %in% c(group1, group2)) %>%
    select(Patient_ID, Cell_type, HLA_score) %>%
    pivot_wider(names_from = Cell_type, values_from = HLA_score)
  
  data1 <- paired_data %>% pull(!!sym(group1))
  data2 <- paired_data %>% pull(!!sym(group2))
  
  test <- t.test(data1, data2, paired = TRUE)
  
  data.frame(
    comparison = paste(group1, "vs", group2),
    group1 = group1,
    group2 = group2,
    n_paired = length(patients_with_both),
    p_value = test$p.value
  )
})

# Print paired results to terminal
cat("\n=== PAIRED T-TEST RESULTS ===\n")
print(paired_results)
cat("\n")

# --- 7. PLOT ---
malignant_colors <- c(
  "NPC-like" = "#003399",
  "OPC-like" = "#6699CC",
  "AC-like" = "#FFB366",
  "GPC-like" = "#FF8899",
  "MES-like" = "#CC0000"
)

p_unpaired <- ggboxplot(
  sig_scores, 
  x = "Cell_type", 
  y = "HLA_score",
  fill = "Cell_type", 
  palette = malignant_colors,
  add = "jitter",
  add.params = list(size = 2, alpha = 0.7),
  outlier.shape = NA
) +
  stat_compare_means(
    comparisons = my_comparisons, 
    method = "t.test",
    paired = FALSE,
    label = "p.format",
    size = 3.5,
    tip.length = 0.01
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10),
    axis.text.y = element_text(color = "black", size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    axis.line = element_line(color = "black", size = 0.5),
    axis.ticks = element_line(color = "black", size = 0.5),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(
    x = "",
    y = "HLA Class I Score"
  )

print(p_unpaired)

# Save to specified directory
ggsave("/n/scratch/users/j/jag9838/GBM_ICB/outputs_final/HLA_Class_I_UNPAIRED_PreICB_upfront.pdf", 
       plot = p_unpaired, width = 6, height = 5, dpi = 300, bg = "white")

# Save results
write.csv(unpaired_results, "unpaired_tests_results_class_I_PreICB_upfront.csv", row.names = FALSE)
write.csv(paired_results, "paired_tests_results_class_I_PreICB_upfront.csv", row.names = FALSE)

cat("=== Analysis Complete ===\n")
cat("Plot saved to: /n/scratch/users/j/jag9838/GBM_ICB/outputs_final/HLA_Class_I_UNPAIRED_PreICB_upfront.pdf\n")
```

# Extended Data Figure 7d

Expression of Verhaak signature genes across patient samples, with corresponding bulk-RNA and snRNA-seq Verhaak annotations

```{r Ex.Fig.7d-Verhaak-classification-expression-by-patient}

pseudobulks_by_sample_pre_TPM <- readRDS(file = "../data/pseudobulks_by_sample_pre_TPM.rds")
pseudobulks_by_sample_cell_type_pre_TPM <- readRDS(file = "../data/pseudobulks_by_sample_cell_type_pre_TPM.rds")
pseudobulks_by_sample_post_TPM <- readRDS(file = "../data/pseudobulks_by_sample_post_TPM.rds")
pseudobulks_by_sample_cell_type_post_TPM <- readRDS(file = "../data/pseudobulks_by_sample_cell_type_post_TPM.rds")

GBM_Verhaak_signatures_df <- readRDS(file = "../data/GBM_Verhaak_signatures_df.rds") 
project_metadata_GBM_verhaak_subtypes <- readRDS(file = "../data/project_metadata_GBM_verhaak_subtypes.rds")



# manually calculate the bulk-RNA state by the maximal GBM cell-state fraction
metadata_malignant <- metadata_ICB %>% dplyr::filter(malignant_group == "Malignant") %>% droplevels()
metadata_malignant %>% dim() # 345094  15
  
MIN_MALIGNANT_SAMPLE_SIZE <- 100

SAMPLES_TO_EXCLUDE <-
  metadata_malignant %>%
  dplyr::group_by(Rx, Patient_ID, Sample_ID) %>%
  dplyr::summarise(total_sample = n(), .groups = "drop") %>%
  # dplyr::arrange(count) %>%
  dplyr::filter(total_sample < MIN_MALIGNANT_SAMPLE_SIZE) %>%
  dplyr::pull(Sample_ID)

SAMPLES_TO_EXCLUDE %>% length() # 3



pseudobulk_verhaak_sample_classifications <-
  cbind(log2(pseudobulks_by_sample_pre_TPM + 1),
        log2(pseudobulks_by_sample_post_TPM + 1)) %>% scalop::rowcenter() %>% 
  scalop::sigScores(sigs = split(GBM_Verhaak_signatures_df$gene_name, GBM_Verhaak_signatures_df$sig_name), expr.center = FALSE) %>% 
  dplyr::mutate(pseudobulk_verhaak_subtype = colnames(.)[max.col(.)]) %>% 
  tibble::rownames_to_column(var = "Sample_ID")



# (B) compare sample-level pseudo-bulks Verhaak mesenchymal expression
pseudobulks_by_sample_pre_TPM %>% dim() # 36601  22

pseudobulks_by_sample_GBM_Verhaak_summary <-
  cbind(log2(pseudobulks_by_sample_pre_TPM[GBM_Verhaak_signatures_df$gene_name, ] + 1),
        log2(pseudobulks_by_sample_post_TPM[GBM_Verhaak_signatures_df$gene_name, ] + 1)) %>% scalop::rowcenter() %>% 
  reshape2::melt(varnames = c("gene_name", "Sample_ID"), value.name = "Exp") %>% 
  dplyr::left_join(GBM_Verhaak_signatures_df, by = "gene_name", relationship = "many-to-many") %>%
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::group_by(Rx, Patient_ID, Sample_ID) %>% 
                     dplyr::summarise(total_sample = n(), .groups = "drop"), by = "Sample_ID") %>% 
  dplyr::left_join(project_metadata %>% 
                     dplyr::select(Patient_ID, Overall_survival_18m), by = "Patient_ID") %>% 
  dplyr::left_join(project_metadata_GBM_verhaak_subtypes %>% 
                     dplyr::select(Patient_ID, Rx, bulk_verhaak_subtype = verhaak_subtype), by = c("Patient_ID", "Rx")) %>% 
  dplyr::left_join(pseudobulk_verhaak_sample_classifications %>% 
                     dplyr::select(Sample_ID, pseudobulk_verhaak_subtype), by = "Sample_ID")



PATIENTS_TO_EXCLUDE <- c()

pseudobulks_by_sample_GBM_Verhaak_summary_filtered <- 
  pseudobulks_by_sample_GBM_Verhaak_summary %>% 
  dplyr::filter(Patient_ID %!in% PATIENTS_TO_EXCLUDE,
                Sample_ID %!in% SAMPLES_TO_EXCLUDE)


# by patient
pseudobulk_verhaak_patient_classifications <- 
  pseudobulk_verhaak_sample_classifications %>% 
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::select(Rx, Patient_ID, Sample_ID) %>% unique(), by = "Sample_ID") %>% 
  dplyr::group_by(Rx, Patient_ID) %>% 
  dplyr::summarise(pseudobulk_verhaak_patient = case_when(n_distinct(pseudobulk_verhaak_subtype) == 1 ~ unique(pseudobulk_verhaak_subtype),
                                                          .default = NA), .groups = "drop")

# pseudobulk_verhaak_patient_classifications %>% print(n = Inf)



pseudobulks_by_patient_GBM_Verhaak_summary_filtered <-
  pseudobulks_by_sample_GBM_Verhaak_summary_filtered %>% 
  dplyr::group_by(Rx, Patient_ID, gene_name, gene_order, sig_name, Overall_survival_18m, bulk_verhaak_subtype) %>% 
  dplyr::summarise(mean_exp_patient = mean(Exp), .groups = "drop") %>% 
  dplyr::left_join(pseudobulk_verhaak_patient_classifications, by = c("Rx", "Patient_ID")) %>% 
  dplyr::mutate(pseudobulk_verhaak_patient_group = case_when(pseudobulk_verhaak_patient == "Mesenchymal" ~ "Mesenchymal",
                                                             .default = "Others") %>% factor(levels = c("Others", "Mesenchymal")))



pseudobulk_patient_order_GBM_Verhaak_MES <-
  pseudobulks_by_patient_GBM_Verhaak_summary_filtered %>% 
  dplyr::filter(sig_name == "Mesenchymal") %>% 
  dplyr::group_by(Rx, Patient_ID, Overall_survival_18m, pseudobulk_verhaak_patient, pseudobulk_verhaak_patient_group) %>% 
  dplyr::summarise(mean_exp = mean(mean_exp_patient), .groups = "drop") %>% 
  arrange(-mean_exp) %>% 
  tibble::rowid_to_column(var = "patient_order")



patient_classification_bulk_verhaak_panel <-
  pseudobulk_patient_order_GBM_Verhaak_MES %>%
  dplyr::mutate(patient_id = case_when(Rx == "Pre-ICB" ~ paste0(" ", Patient_ID), # to display the same patient from different timepoints
                                       .default = Patient_ID)) %>% 
  dplyr::left_join(project_metadata_GBM_verhaak_subtypes %>%
                     dplyr::select(Patient_ID, Rx, bulk_verhaak_subtype = verhaak_subtype), by = c("Patient_ID", "Rx")) %>%
  ggplot(aes(x= reorder(patient_id, -patient_order), y = "Verhhak GBM state (bulk)")) +
  geom_tile(aes(fill = bulk_verhaak_subtype), alpha = 0.8, show.legend = FALSE) +
  facet_grid(cols = vars(pseudobulk_verhaak_patient_group), scales = "free", space = "free") +
  scale_fill_manual(values = c(VERHAAK_SUBTYPE_CLASSICAL_C, VERHAAK_SUBTYPE_MESENCHYMAL_C, VERHAAK_SUBTYPE_NEURAL_C, VERHAAK_SUBTYPE_PRONEURAL_C), na.value = "white") +
  scale_x_discrete(expand = c(0,0)) + scale_y_discrete(expand = c(0,0)) +
  labs(x = "", y = "", fill = "Verhaak GBM state") +
  default_theme_xrot_90_border() +
  theme(legend.key.spacing.y = unit(10,"pt"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())


patient_classification_pseudobulk_verhaak_panel <-
  pseudobulk_patient_order_GBM_Verhaak_MES %>%
  dplyr::mutate(patient_id = case_when(Rx == "Pre-ICB" ~ paste0(" ", Patient_ID), # to display the same patient from different timepoints
                                       .default = Patient_ID)) %>% 
  dplyr::left_join(project_metadata_GBM_verhaak_subtypes %>%
                     dplyr::select(Patient_ID, Rx, bulk_verhaak_subtype = verhaak_subtype), by = c("Patient_ID", "Rx")) %>%
  ggplot(aes(x= reorder(patient_id, -patient_order), y = "Verhhak GBM state (pseudobulk)")) +
  geom_tile(aes(fill = pseudobulk_verhaak_patient), alpha = 0.8) +
  facet_grid(cols = vars(pseudobulk_verhaak_patient_group), scales = "free", space = "free") +
  scale_fill_manual(values = c(VERHAAK_SUBTYPE_CLASSICAL_C, VERHAAK_SUBTYPE_MESENCHYMAL_C, VERHAAK_SUBTYPE_NEURAL_C, VERHAAK_SUBTYPE_PRONEURAL_C), na.value = "white") +
  scale_x_discrete(expand = c(0,0)) + scale_y_discrete(expand = c(0,0)) +
  labs(x = "", y = "", fill = "Verhaak GBM state (pseudobulk)") +
  default_theme_xrot_90_border() +
  theme(strip.background.x = element_blank(), strip.text.x = element_blank(),
        legend.key.spacing.y = unit(10,"pt"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())



overall_survival_18m_patient_panel <-
  pseudobulk_patient_order_GBM_Verhaak_MES %>% 
  dplyr::mutate(patient_id = case_when(Rx == "Pre-ICB" ~ paste0(" ", Patient_ID), # to display the same patient from different timepoints
                                       .default = Patient_ID)) %>% 
  dplyr::left_join(pseudobulk_verhaak_patient_classifications, by = c("Rx", "Patient_ID")) %>% 
  ggplot(aes(x= reorder(patient_id, -patient_order), y = "Overall survival")) + 
  geom_tile(aes(fill = Overall_survival_18m), alpha = 0.8) + 
  facet_grid(cols = vars(pseudobulk_verhaak_patient_group), scales = "free", space = "free") + 
  scale_fill_manual(values = c(SURVIVAL_SS_C, SURVIVAL_LS_C), na.value = "white") + 
  scale_x_discrete(expand = c(0,0)) + scale_y_discrete(expand = c(0,0)) + 
  labs(x = "", y = "", fill = "Overall survival") + 
  default_theme_xrot_90_border() +
  theme(strip.background.x = element_blank(), strip.text.x = element_blank(),
        legend.key.spacing.y = unit(10,"pt"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
  


patient_RX_ICB_timepoint_panel <-
  pseudobulk_patient_order_GBM_Verhaak_MES %>% 
  dplyr::mutate(patient_id = case_when(Rx == "Pre-ICB" ~ paste0(" ", Patient_ID), # to display the same patient from different timepoints
                                       .default = Patient_ID)) %>% 
  ggplot(aes(x= reorder(patient_id, -patient_order), y = "Timepoint")) + 
  geom_tile(aes(fill = Rx), alpha = 0.8) + 
  facet_grid(cols = vars(pseudobulk_verhaak_patient_group), scales = "free", space = "free") + 
  scale_fill_manual(values = c(RX_PRE_ICB_C, RX_POST_ICB_C)) + 
  scale_x_discrete(expand = c(0,0)) + scale_y_discrete(expand = c(0,0)) + 
  labs(x = "", y = "", fill = "") + 
  default_theme_xrot_90_border() +
  theme(strip.background.x = element_blank(), strip.text.x = element_blank(),
        legend.key.spacing.y = unit(10,"pt"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())



MAX_LOG2_EXPRESSION_THRESHOLD <- 4

pseudobulks_by_patient_GBM_Verhaak_heatmap <- 
  pseudobulks_by_patient_GBM_Verhaak_summary_filtered %>% 
  dplyr::left_join(pseudobulk_patient_order_GBM_Verhaak_MES %>% 
                     dplyr::select(Rx, Patient_ID, patient_order), by = c("Rx", "Patient_ID")) %>% 
  dplyr::mutate(mean_exp_patient_trim = case_when(mean_exp_patient < -MAX_LOG2_EXPRESSION_THRESHOLD ~ -MAX_LOG2_EXPRESSION_THRESHOLD,
                                                  MAX_LOG2_EXPRESSION_THRESHOLD < mean_exp_patient ~ MAX_LOG2_EXPRESSION_THRESHOLD,
                                                  .default = mean_exp_patient)) %>% 
  dplyr::mutate(patient_id = case_when(Rx == "Pre-ICB" ~ paste0(" ", Patient_ID), # to display the same patient from different timepoints
                                       .default = Patient_ID)) %>% 
  ggplot(aes(x= reorder(patient_id, -patient_order), y = reorder(gene_name, -gene_order))) + 
  geom_tile(aes(fill = mean_exp_patient_trim)) + 
  facet_grid(cols = vars(pseudobulk_verhaak_patient_group), rows = vars(sig_name), scales = "free", space = "free") + 
  scale_fill_gradient2(low = HEATMAP_LOW_VALUE_C, high = HEATMAP_HIGH_VALUE_C, mid = "white", midpoint = 0,
                       guide = guide_colorbar(frame.colour = "black", ticks.colour = "black", frame.linewidth = 0.5)) + 
  labs(x = "", y = "Verhaak signatures genes", fill = "Exp, log2") + 
  default_theme_xrot_90_border() +
  theme(strip.background.x = element_blank(), strip.text.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank())



(patient_classification_bulk_verhaak_panel / plot_spacer() / patient_classification_pseudobulk_verhaak_panel / plot_spacer() / overall_survival_18m_patient_panel / plot_spacer() / patient_RX_ICB_timepoint_panel / plot_spacer() /  pseudobulks_by_patient_GBM_Verhaak_heatmap) + 
  patchwork::plot_layout(heights = c(1,-1.4,1,-1.4,1,-1.4,1,-1.4,25), 
                         guides = "collect") +
  patchwork::plot_annotation(subtitle = ifelse(length(PATIENTS_TO_EXCLUDE) == 0, 
                                               yes = "", no = paste0("Excluded patients: ", paste(PATIENTS_TO_EXCLUDE, collapse = ", "))), 
                             theme = default_theme())

```

# Extended Data Figure 8a

```{r annotate-samples-by-Verhaak-scores-pseudobulk}

pseudobulks_by_sample_pre_TPM <- readRDS(file = "../data/pseudobulks_by_sample_pre_TPM.rds")
pseudobulks_by_sample_cell_type_pre_TPM <- readRDS(file = "../data/pseudobulks_by_sample_cell_type_pre_TPM.rds")
pseudobulks_by_sample_post_TPM <- readRDS(file = "../data/pseudobulks_by_sample_post_TPM.rds")
pseudobulks_by_sample_cell_type_post_TPM <- readRDS(file = "../data/pseudobulks_by_sample_cell_type_post_TPM.rds")

GBM_Verhaak_signatures_df <- readRDS(file = "../data/GBM_Verhaak_signatures_df.rds") 
project_metadata_GBM_verhaak_subtypes <- readRDS(file = "../data/project_metadata_GBM_verhaak_subtypes.rds")


pseudobulk_verhaak_sample_classifications <-
  cbind(log2(pseudobulks_by_sample_pre_TPM + 1),
        log2(pseudobulks_by_sample_post_TPM + 1)) %>% scalop::rowcenter() %>% 
  scalop::sigScores(sigs = split(GBM_Verhaak_signatures_df$gene_name, GBM_Verhaak_signatures_df$sig_name), expr.center = FALSE) %>% 
  dplyr::mutate(pseudobulk_verhaak_subtype = colnames(.)[max.col(.)]) %>% 
  tibble::rownames_to_column(var = "Sample_ID")



pseudobulk_verhaak_patient_classifications <- 
  pseudobulk_verhaak_sample_classifications %>% 
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::select(Rx, Patient_ID, Sample_ID) %>% unique(), by = "Sample_ID") %>% 
  dplyr::group_by(Rx, Patient_ID) %>% 
  dplyr::summarise(pseudobulk_verhaak_patient = case_when(n_distinct(pseudobulk_verhaak_subtype) == 1 ~ unique(pseudobulk_verhaak_subtype),
                                                          .default = NA), .groups = "drop")

```

```{r Ex.Fig.8a-relative-fraction-of-MESENCHYMAL-signal-by-cell-type}

# Load TPM expression matrix to filter genes
exp_mat_all_tpm_HLA <- readRDS("../data/exp_mat_all_tpm_HLA.rds")


# ============================================================================
# ANALYSIS: Fraction of MES signal by cell type 
# ============================================================================

# (1) Calculate total MES reads by sample and cell type
total_mesanchymal_reads_by_sample_cell_type_summary <-
  pseudobulks_by_sample_cell_type_pre_TPM[GBM_Verhaak_signatures_df %>% 
                                             dplyr::filter(sig_name == "Mesenchymal") %>% 
                                             dplyr::pull(gene_name), ] %>% 
  colSums() %>% 
  data.frame(Verhaak_MES_reads = .) %>% 
  tibble::rownames_to_column(var = "sample_cell_type_id") %>%
  tidyr::separate(col = "sample_cell_type_id", into = c("Sample_ID", "Cell_type"), sep = "[*]") %>%
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::filter(Rx == "Pre-ICB") %>% 
                     dplyr::select(Sample_ID, Patient_ID, Rx) %>% 
                     unique(), by = "Sample_ID") %>% 
  dplyr::left_join(pseudobulk_verhaak_patient_classifications_pre %>% 
                     dplyr::select(Patient_ID, pseudobulk_verhaak_patient), by = "Patient_ID") %>% 
  dplyr::filter(!is.na(pseudobulk_verhaak_patient))


# (2) Combine all malignant cell types into one "Malignant" group
total_mesanchymal_reads_combined_malignant <-
  total_mesanchymal_reads_by_sample_cell_type_summary %>%
  dplyr::mutate(Cell_type_combined = case_when(
    Cell_type %in% c("AC-like", "MES-like", "GPC-like", "NPC-like", "OPC-like", "Malignant (ND)") ~ "Malignant",
    .default = Cell_type
  )) %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(Verhaak_MES_reads = sum(Verhaak_MES_reads), .groups = "drop")


# (3) Calculate fraction of MES signal by cell type at sample level
fraction_of_mes_signal_by_sample <-
  total_mesanchymal_reads_combined_malignant %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient) %>%
  dplyr::mutate(total_sample_MES_reads = sum(Verhaak_MES_reads),
                fraction_MES_signal = Verhaak_MES_reads / total_sample_MES_reads)

# (4) Average to patient level (Daniel's approach - averaging across multiple samples per patient)
fraction_of_mes_signal_by_patient <-
  fraction_of_mes_signal_by_sample %>%
  dplyr::group_by(Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(mean_fraction_MES_signal = mean(fraction_MES_signal), .groups = "drop")

# (5) Filter to only Mesenchymal tumors
fraction_of_mes_signal_by_patient_MES_only <-
  fraction_of_mes_signal_by_patient %>%
  dplyr::filter(pseudobulk_verhaak_patient == "Mesenchymal")

# NEW: Complete the data to include all combinations with zeros for missing cell types
# Get all unique patients
all_patients <- fraction_of_mes_signal_by_patient_MES_only %>%
  dplyr::select(Patient_ID, Rx, pseudobulk_verhaak_patient) %>%
  unique()

# Get all unique cell types
all_cell_types <- fraction_of_mes_signal_by_patient_MES_only %>%
  dplyr::select(Cell_type_combined) %>%
  unique()

# Create complete grid and fill missing with 0
fraction_of_mes_signal_by_patient_MES_only <-
  all_patients %>%
  tidyr::crossing(all_cell_types) %>%
  dplyr::left_join(fraction_of_mes_signal_by_patient_MES_only, 
                   by = c("Patient_ID", "Rx", "pseudobulk_verhaak_patient", "Cell_type_combined")) %>%
  dplyr::mutate(mean_fraction_MES_signal = 
                  tidyr::replace_na(mean_fraction_MES_signal, 0))


# (6) Order cell types by median fraction in MES tumors
cell_type_order <-
  fraction_of_mes_signal_by_patient_MES_only %>%
  dplyr::group_by(Cell_type_combined) %>%
  dplyr::summarise(median_fraction = median(mean_fraction_MES_signal), .groups = "drop") %>%
  dplyr::arrange(-median_fraction) %>%
  tibble::rowid_to_column(var = "order")

# (7) Create the plot
fraction_of_mes_signal_by_patient_MES_only %>%
  dplyr::left_join(cell_type_order %>% 
                     dplyr::select(Cell_type_combined, order), 
                   by = "Cell_type_combined") %>%
  ggplot(aes(x = reorder(Cell_type_combined, order), 
             y = mean_fraction_MES_signal)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.8, 
               fill = "#E41A1C") +
  geom_point(position = position_jitter(width = 0.2), 
             size = 2, alpha = 0.5) +
  scale_y_continuous(labels = scales::percent_format(), 
                     limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "", 
       y = "Fraction of Total MES Signal",
       title = "Contribution to Verhaak MES Signature by Cell Type (Pre-ICB)",
       subtitle = "Mesenchymal tumors only (samples with >100 total cells, absent cell types shown as 0%)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10))

```

# Extended Data Figure 8b

left panel: Compare the source of the HLA class I signal by cell type (as the fraction of the total HLA class I signal in a given sample) in MESENCHYMAL vs. non-MESENCHYMAL samples

```{r Ex.Fig.8b-left-HLA-class-I-signal-by-cell-type-in-MES-vs-non-MES-samples}

# Load TPM expression matrix to filter genes
exp_mat_all_tpm_HLA <- readRDS("../data/exp_mat_all_tpm_HLA.rds")


# ============================================================================
# CREATE HLA CLASS I SIGNATURE
# ============================================================================

HLA_CLASS_I_GENES <- c("B2M", "HLA-A", "HLA-B", "HLA-C", "HLA-E", "HLA-F", "HLA-G", "HLA-H")

# Create signature list
HLA_signatures <- list("HLA_Class_I" = HLA_CLASS_I_GENES)

# Filter to genes that exist in your dataset
HLA_signatures_clean <- lapply(HLA_signatures, function(sig_genes) { 
  sig_genes[sig_genes %in% rownames(exp_mat_all_tpm_HLA)]
})

# Print how many genes we have per signature
print("Number of genes per signature:")
print(sapply(HLA_signatures_clean, length))

# Create the dataframe
HLA_signatures_df <- 
  map(names(HLA_signatures_clean), function(HLA_signature) {
    HLA_signatures_clean[[HLA_signature]] %>% 
      data.frame(gene_name = .) %>% 
      dplyr::mutate(sig_name = HLA_signature)
  }) %>% 
  do.call(rbind, .) %>% 
  tibble::rowid_to_column(var = "gene_order")


# ============================================================================
# ANALYSIS: Fraction of HLA Class I signal by cell type (Daniel's approach)
# ============================================================================

# (1) Calculate total HLA Class I reads by sample and cell type
total_hla_class_i_reads_by_sample_cell_type_summary <-
  pseudobulks_by_sample_cell_type_pre_TPM[HLA_signatures_df %>% 
                                            dplyr::filter(sig_name == "HLA_Class_I") %>% 
                                            dplyr::pull(gene_name), ] %>% 
  colSums() %>% 
  data.frame(HLA_Class_I_reads = .) %>% 
  tibble::rownames_to_column(var = "sample_cell_type_id") %>%
  tidyr::separate(col = "sample_cell_type_id", into = c("Sample_ID", "Cell_type"), sep = "[*]") %>%
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::filter(Rx == "Pre-ICB") %>% 
                     dplyr::select(Sample_ID, Patient_ID, Rx) %>% 
                     unique(), by = "Sample_ID") %>% 
  dplyr::left_join(pseudobulk_verhaak_patient_classifications %>% 
                     dplyr::filter(Rx == "Pre-ICB") %>% 
                     dplyr::select(Patient_ID, pseudobulk_verhaak_patient) %>% unique(), by = "Patient_ID") %>% 
  dplyr::filter(!is.na(pseudobulk_verhaak_patient))


# (2) Combine all malignant cell types into one "Malignant" group
total_hla_class_i_reads_combined_malignant <-
  total_hla_class_i_reads_by_sample_cell_type_summary %>%
  dplyr::mutate(Cell_type_combined = case_when(
    Cell_type %in% c("AC-like", "MES-like", "GPC-like", "NPC-like", "OPC-like", "Malignant (ND)") ~ "Malignant",
    .default = Cell_type
  )) %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(HLA_Class_I_reads = sum(HLA_Class_I_reads), .groups = "drop")

# (3) Calculate fraction of HLA Class I signal by cell type at sample level (Daniel's approach)
fraction_of_hla_class_i_signal_by_sample <-
  total_hla_class_i_reads_combined_malignant %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient) %>%
  dplyr::mutate(total_sample_HLA_Class_I_reads = sum(HLA_Class_I_reads),
                fraction_HLA_Class_I_signal = HLA_Class_I_reads / total_sample_HLA_Class_I_reads)

# (4) Average to patient level (Daniel's approach - averaging across multiple samples per patient)
fraction_of_hla_class_i_signal_by_patient <-
  fraction_of_hla_class_i_signal_by_sample %>%
  dplyr::group_by(Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(mean_fraction_HLA_Class_I_signal = mean(fraction_HLA_Class_I_signal), .groups = "drop")

# (5) Create tumor subtype grouping
fraction_of_hla_class_i_signal_by_patient <-
  fraction_of_hla_class_i_signal_by_patient %>%
  dplyr::mutate(Tumor_Subtype = case_when(
    pseudobulk_verhaak_patient == "Mesenchymal" ~ "Mesenchymal",
    .default = "Non-Mesenchymal"
  ))


# NEW: Complete the data to include all combinations with zeros for missing cell types
# Get all unique patients per tumor subtype
all_patients <- fraction_of_hla_class_i_signal_by_patient %>%
  dplyr::select(Patient_ID, Rx, pseudobulk_verhaak_patient, Tumor_Subtype) %>%
  unique()

# Get all unique cell types
all_cell_types <- fraction_of_hla_class_i_signal_by_patient %>%
  dplyr::select(Cell_type_combined) %>%
  unique()

# Create complete grid and fill missing with 0
fraction_of_hla_class_i_signal_by_patient <-
  all_patients %>%
  tidyr::crossing(all_cell_types) %>%
  dplyr::left_join(fraction_of_hla_class_i_signal_by_patient, 
                   by = c("Patient_ID", "Rx", "pseudobulk_verhaak_patient", 
                          "Tumor_Subtype", "Cell_type_combined")) %>%
  dplyr::mutate(mean_fraction_HLA_Class_I_signal = 
                  tidyr::replace_na(mean_fraction_HLA_Class_I_signal, 0))


# (6) Order cell types by median fraction in MES tumors
cell_type_order <-
  fraction_of_hla_class_i_signal_by_patient %>%
  dplyr::filter(Tumor_Subtype == "Mesenchymal") %>%
  dplyr::group_by(Cell_type_combined) %>%
  dplyr::summarise(median_fraction = median(mean_fraction_HLA_Class_I_signal), .groups = "drop") %>%
  dplyr::arrange(-median_fraction) %>%
  tibble::rowid_to_column(var = "order")

# (7) Create the plot
fraction_of_hla_class_i_signal_by_patient %>%
  dplyr::left_join(cell_type_order %>% 
                     dplyr::select(Cell_type_combined, order), 
                   by = "Cell_type_combined") %>%
  ggplot(aes(x = reorder(Cell_type_combined, order), 
             y = mean_fraction_HLA_Class_I_signal, 
             fill = Tumor_Subtype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.8, 
               position = position_dodge(width = 0.8)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), 
             size = 2, alpha = 0.5) +
  scale_fill_manual(values = c("Mesenchymal" = "#E41A1C", 
                               "Non-Mesenchymal" = "#377EB8")) +
  scale_y_continuous(labels = scales::percent_format(), 
                     limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "", 
       y = "Fraction of Total HLA Class I Signal",
       fill = "Tumor Subtype",
       title = "Contribution to HLA Class I Signature by Cell Type (Pre-ICB)",
       subtitle = "Samples with >100 total cells") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 12),
        legend.position = "top",
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10))

```

# Extended Data Figure 8c

Compare HLA class II scores, based on bulk expression scores, per malignant cell-state

```{r Ex.Fig.8c-compare-bulk-expression-HLA-class-II-scores-by-malignant-cell-state}

exp_mat_all_tpm_HLA <- readRDS("backup_temp_data/Single cell portal/exp_mat_all_tpm_HLA.rds")


HLA_CLASS_II_GENES <- c("HLA-DMA", "HLA-DMB", "HLA-DOA", "HLA-DOB", "HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DQA2", "HLA-DQB1", "HLA-DQB2", "HLA-DRA", "HLA-DRB1", "HLA-DRB3", "HLA-DRB4", "HLA-DRB5")

genes_present <- HLA_CLASS_II_GENES[HLA_CLASS_II_GENES %in% rownames(exp_mat_all_tpm_HLA)]

MIN_NUMBER_OF_CELL_PER_CELL_TYPE <- 5

# FILTER TO PRE-ICB SAMPLES IMMEDIATELY
sample_cell_type_counts_summary <- 
  metadata_ICB %>%
  dplyr::filter(Rx == "Pre-ICB") %>%  # <- FILTER UPFRONT HERE
  group_by(Rx, Patient_ID, Sample_ID, malignant_group, Cell_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  dplyr::filter(count >= MIN_NUMBER_OF_CELL_PER_CELL_TYPE)

# Calculate pseudobulk for ALL malignant cell types (including "Malignant (ND)")
# Now only using Pre-ICB samples
bulk_expression_by_cell_type_patient_malignant <- 
  sample_cell_type_counts_summary %>%
  dplyr::filter(malignant_group == "Malignant") %>%
  droplevels() %>%
  pull(Cell_type) %>% 
  levels() %>%
  map(function(cur_cell_type) {
    
    cat("Processing:", cur_cell_type, "\n")
    
    cur_cell_type_counts_summary <- sample_cell_type_counts_summary %>%
      dplyr::filter(Cell_type == cur_cell_type) %>% 
      droplevels()
    
    if (nrow(cur_cell_type_counts_summary) == 0) {
      return(NULL)
    }
    
    bulk_expression_cur_cell_type <- 
      cur_cell_type_counts_summary$Sample_ID %>% unique() %>%
      map(function(cur_cell_type_sample) {
        
        cur_cell_type_sample_cell_id <- metadata_ICB %>%
          dplyr::filter(Cell_type == cur_cell_type, Sample_ID == cur_cell_type_sample) %>%
          pull(Cell_ID)
        
        exp_subset <- exp_mat_all_tpm_HLA[, cur_cell_type_sample_cell_id, drop = FALSE]
        log2(Matrix::rowMeans(exp_subset) + 1)
        
      }) %>% do.call(cbind, .)
    
    colnames(bulk_expression_cur_cell_type) <- cur_cell_type_counts_summary$Sample_ID %>% unique()
    
    bulk_expression_cur_cell_type %>%
      reshape2::melt(varnames = c("gene_name", "Sample_ID"), value.name = "Exp") %>%
      left_join(sample_cell_type_counts_summary %>%
                  dplyr::filter(Cell_type == cur_cell_type) %>% 
                  droplevels() %>%
                  dplyr::select(Rx, Patient_ID, Sample_ID) %>% 
                  unique(), 
                by = "Sample_ID") %>%
      group_by(Rx, Patient_ID, gene_name) %>%
      summarise(mean_exp = mean(Exp), .groups = "drop") %>%
      mutate(Cell_type = cur_cell_type)
    
  }) %>% 
  keep(~ !is.null(.)) %>%
  do.call(rbind, .) %>%
  mutate_at(c("gene_name"), as.character)

# EXCLUDE "Malignant (ND)" BEFORE centering - USING scalop::rowcenter()
# Now centering only across Pre-ICB samples
bulk_expression_centered_matrix <- 
  bulk_expression_by_cell_type_patient_malignant %>%
  dplyr::filter(Cell_type != "Malignant (ND)") %>%  # FILTER BEFORE CENTERING
  mutate(rx_patient_cell_type_id = paste0(Rx, "*", Patient_ID, "*", Cell_type)) %>%
  pivot_wider(id_cols = c("gene_name"), 
              names_from = "rx_patient_cell_type_id", 
              values_from = "mean_exp") %>%
  column_to_rownames(var = "gene_name") %>%
  as.matrix() %>%
  scalop::rowcenter()  # CENTERING ONLY PRE-ICB DATA

# Calculate scores
sig_scores <- bulk_expression_centered_matrix %>%
  as.data.frame() %>%
  rownames_to_column("gene_name") %>%
  dplyr::filter(gene_name %in% genes_present) %>%
  pivot_longer(-gene_name, names_to = "id", values_to = "centered_exp") %>%
  group_by(id) %>%
  summarise(HLA_score = mean(centered_exp, na.rm = TRUE), .groups = "drop") %>%
  separate(id, into = c("Rx", "Patient_ID", "Cell_type"), sep = "\\*")
# NO FILTER HERE - already only Pre-ICB data

# Order cell types by median HLA score (highest to lowest)
cell_type_order <- sig_scores %>%
  group_by(Cell_type) %>%
  summarise(median_score = median(HLA_score, na.rm = TRUE)) %>%
  arrange(desc(median_score)) %>%
  pull(Cell_type)

sig_scores <- sig_scores %>%
  mutate(Cell_type = factor(Cell_type, levels = cell_type_order))

# --- 6. STATISTICS ---
my_comparisons <- list(
  c("MES-like", "AC-like"),
  c("MES-like", "GPC-like"),
  c("MES-like", "NPC-like"),
  c("MES-like", "OPC-like")
)

unpaired_results <- map_dfr(my_comparisons, function(comp) {
  group1 <- comp[1]
  group2 <- comp[2]
  
  data1 <- sig_scores %>% filter(Cell_type == group1) %>% pull(HLA_score)
  data2 <- sig_scores %>% filter(Cell_type == group2) %>% pull(HLA_score)
  
  test <- t.test(data1, data2, paired = FALSE)
  
  data.frame(
    comparison = paste(group1, "vs", group2),
    group1 = group1,
    group2 = group2,
    n1 = length(data1),
    n2 = length(data2),
    p_value = test$p.value
  )
})

paired_results <- map_dfr(my_comparisons, function(comp) {
  group1 <- comp[1]
  group2 <- comp[2]
  
  patients_with_both <- sig_scores %>%
    filter(Cell_type %in% c(group1, group2)) %>%
    group_by(Patient_ID) %>%
    summarise(n_types = n_distinct(Cell_type), .groups = "drop") %>%
    filter(n_types == 2) %>%
    pull(Patient_ID)
  
  if (length(patients_with_both) < 3) {
    return(data.frame(
      comparison = paste(group1, "vs", group2),
      group1 = group1,
      group2 = group2,
      n_paired = length(patients_with_both),
      p_value = NA
    ))
  }
  
  paired_data <- sig_scores %>%
    filter(Patient_ID %in% patients_with_both, Cell_type %in% c(group1, group2)) %>%
    select(Patient_ID, Cell_type, HLA_score) %>%
    pivot_wider(names_from = Cell_type, values_from = HLA_score)
  
  data1 <- paired_data %>% pull(!!sym(group1))
  data2 <- paired_data %>% pull(!!sym(group2))
  
  test <- t.test(data1, data2, paired = TRUE)
  
  data.frame(
    comparison = paste(group1, "vs", group2),
    group1 = group1,
    group2 = group2,
    n_paired = length(patients_with_both),
    p_value = test$p.value
  )
})

# Print paired results to terminal
cat("\n=== PAIRED T-TEST RESULTS ===\n")
print(paired_results)
cat("\n")

# --- 7. PLOT ---
malignant_colors <- c(
  "NPC-like" = "#003399",
  "OPC-like" = "#6699CC",
  "AC-like" = "#FFB366",
  "GPC-like" = "#FF8899",
  "MES-like" = "#CC0000"
)

p_unpaired <- ggboxplot(
  sig_scores, 
  x = "Cell_type", 
  y = "HLA_score",
  fill = "Cell_type", 
  palette = malignant_colors,
  add = "jitter",
  add.params = list(size = 2, alpha = 0.7),
  outlier.shape = NA
) +
  stat_compare_means(
    comparisons = my_comparisons, 
    method = "t.test",
    paired = FALSE,
    label = "p.format",
    size = 3.5,
    tip.length = 0.01
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10),
    axis.text.y = element_text(color = "black", size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    axis.line = element_line(color = "black", size = 0.5),
    axis.ticks = element_line(color = "black", size = 0.5),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
  labs(
    x = "",
    y = "HLA Class II Score"
  )

print(p_unpaired)

```

# Extended Data Figure 8d

left panel: Compare the source of the HLA class II signal by cell type (as the fraction of the total HLA class II signal in a given sample) in MESENCHYMAL vs. non-MESENCHYMAL samples

```{r Ex.Fig.8d-left-HLA-class-II-signal-by-cell-type-in-MES-vs-non-MES-samples}

# Load TPM expression matrix to filter genes
exp_mat_all_tpm_HLA <- readRDS("../data/exp_mat_all_tpm_HLA.rds")


# ============================================================================
# CREATE HLA CLASS II SIGNATURE
# ============================================================================

HLA_CLASS_II_GENES <- c("B2M", "CD74", "HLA-DMA", "HLA-DMB", "HLA-DOA", "HLA-DOB", 
                        "HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DQA2", "HLA-DQB1", 
                        "HLA-DQB2", "HLA-DRA", "HLA-DRB1", "HLA-DRB3", "HLA-DRB4", "HLA-DRB5")

# Create signature list
HLA_signatures <- list("HLA_Class_II" = HLA_CLASS_II_GENES)

# Filter to genes that exist in your dataset
HLA_signatures_clean <- lapply(HLA_signatures, function(sig_genes) { 
  sig_genes[sig_genes %in% rownames(exp_mat_all_tpm_HLA)]
})

# Print how many genes we have per signature
print("Number of genes per signature:")
print(sapply(HLA_signatures_clean, length))

# Create the dataframe
HLA_signatures_df <- 
  map(names(HLA_signatures_clean), function(HLA_signature) {
    HLA_signatures_clean[[HLA_signature]] %>% 
      data.frame(gene_name = .) %>% 
      dplyr::mutate(sig_name = HLA_signature)
  }) %>% 
  do.call(rbind, .) %>% 
  tibble::rowid_to_column(var = "gene_order")


# ============================================================================
# ANALYSIS: Fraction of HLA Class II signal by cell type (Daniel's approach)
# ============================================================================

# (1) Calculate total HLA Class II reads by sample and cell type
total_hla_class_ii_reads_by_sample_cell_type_summary <-
  pseudobulks_by_sample_cell_type_pre_TPM[HLA_signatures_df %>% 
                                            dplyr::filter(sig_name == "HLA_Class_II") %>% 
                                            dplyr::pull(gene_name), ] %>% 
  colSums() %>% 
  data.frame(HLA_Class_II_reads = .) %>% 
  tibble::rownames_to_column(var = "sample_cell_type_id") %>%
  tidyr::separate(col = "sample_cell_type_id", into = c("Sample_ID", "Cell_type"), sep = "[*]") %>%
  dplyr::left_join(metadata_ICB %>% 
                     dplyr::filter(Rx == "Pre-ICB") %>% 
                     dplyr::select(Sample_ID, Patient_ID, Rx) %>% 
                     unique(), by = "Sample_ID") %>% 
  dplyr::left_join(pseudobulk_verhaak_patient_classifications %>% 
                     dplyr::filter(Rx == "Pre-ICB") %>% 
                     dplyr::select(Patient_ID, pseudobulk_verhaak_patient) %>% unique(), by = "Patient_ID") %>% 
  dplyr::filter(!is.na(pseudobulk_verhaak_patient))


# (2) Combine all malignant cell types into one "Malignant" group
total_hla_class_ii_reads_combined_malignant <-
  total_hla_class_ii_reads_by_sample_cell_type_summary %>%
  dplyr::mutate(Cell_type_combined = case_when(
    Cell_type %in% c("AC-like", "MES-like", "GPC-like", "NPC-like", "OPC-like", "Malignant (ND)") ~ "Malignant",
    .default = Cell_type
  )) %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(HLA_Class_II_reads = sum(HLA_Class_II_reads), .groups = "drop")

# (3) Calculate fraction of HLA Class I signal by cell type at sample level (Daniel's approach)
fraction_of_hla_class_ii_signal_by_sample <-
  total_hla_class_ii_reads_combined_malignant %>%
  dplyr::group_by(Sample_ID, Patient_ID, Rx, pseudobulk_verhaak_patient) %>%
  dplyr::mutate(total_sample_HLA_Class_II_reads = sum(HLA_Class_II_reads),
                fraction_HLA_Class_II_signal = HLA_Class_II_reads / total_sample_HLA_Class_II_reads)

# (4) Average to patient level (Daniel's approach - averaging across multiple samples per patient)
fraction_of_hla_class_ii_signal_by_patient <-
  fraction_of_hla_class_ii_signal_by_sample %>%
  dplyr::group_by(Patient_ID, Rx, pseudobulk_verhaak_patient, Cell_type_combined) %>%
  dplyr::summarise(mean_fraction_HLA_Class_II_signal = mean(fraction_HLA_Class_II_signal), .groups = "drop")

# (5) Create tumor subtype grouping
fraction_of_hla_class_ii_signal_by_patient <-
  fraction_of_hla_class_ii_signal_by_patient %>%
  dplyr::mutate(Tumor_Subtype = case_when(
    pseudobulk_verhaak_patient == "Mesenchymal" ~ "Mesenchymal",
    .default = "Non-Mesenchymal"
  ))


# NEW: Complete the data to include all combinations with zeros for missing cell types
# Get all unique patients per tumor subtype
all_patients <- fraction_of_hla_class_ii_signal_by_patient %>%
  dplyr::select(Patient_ID, Rx, pseudobulk_verhaak_patient, Tumor_Subtype) %>%
  unique()

# Get all unique cell types
all_cell_types <- fraction_of_hla_class_ii_signal_by_patient %>%
  dplyr::select(Cell_type_combined) %>%
  unique()

# Create complete grid and fill missing with 0
fraction_of_hla_class_ii_signal_by_patient <-
  all_patients %>%
  tidyr::crossing(all_cell_types) %>%
  dplyr::left_join(fraction_of_hla_class_ii_signal_by_patient, 
                   by = c("Patient_ID", "Rx", "pseudobulk_verhaak_patient", 
                          "Tumor_Subtype", "Cell_type_combined")) %>%
  dplyr::mutate(mean_fraction_HLA_Class_II_signal = 
                  tidyr::replace_na(mean_fraction_HLA_Class_II_signal, 0))


# (6) Order cell types by median fraction in MES tumors
cell_type_order <-
  fraction_of_hla_class_ii_signal_by_patient %>%
  dplyr::filter(Tumor_Subtype == "Mesenchymal") %>%
  dplyr::group_by(Cell_type_combined) %>%
  dplyr::summarise(median_fraction = median(mean_fraction_HLA_Class_II_signal), .groups = "drop") %>%
  dplyr::arrange(-median_fraction) %>%
  tibble::rowid_to_column(var = "order")

# (7) Create the plot
fraction_of_hla_class_ii_signal_by_patient %>%
  dplyr::left_join(cell_type_order %>% 
                     dplyr::select(Cell_type_combined, order), 
                   by = "Cell_type_combined") %>%
  ggplot(aes(x = reorder(Cell_type_combined, order), 
             y = mean_fraction_HLA_Class_II_signal, 
             fill = Tumor_Subtype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, linewidth = 0.8, 
               position = position_dodge(width = 0.8)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), 
             size = 2, alpha = 0.5) +
  scale_fill_manual(values = c("Mesenchymal" = "#E41A1C", 
                               "Non-Mesenchymal" = "#377EB8")) +
  scale_y_continuous(labels = scales::percent_format(), 
                     limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "", 
       y = "Fraction of Total HLA Class II Signal",
       fill = "Tumor Subtype",
       title = "Contribution to HLA Class II Signature by Cell Type (Pre-ICB)",
       subtitle = "Samples with >100 total cells") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 12),
        legend.position = "top",
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10))

```

# Extended Data Figure 8e

Relative expression of the HLA genes (class I and II) by malignant cell-state level. Expression was calculated based on bulk expression (averaging over malignant-state cells)

```{r Ex.Fig.8e-relative-expression-of-HLA-genes-by-malignant-state}

library(ComplexHeatmap)
library(circlize)
library(scalop) 


exp_mat_all_tpm_HLA <- readRDS("../data/exp_mat_all_tpm_HLA.rds")


# Keep B2M in BOTH gene lists for calculation
HLA_CLASS_I_GENES <- c("B2M", "HLA-A", "HLA-B", "HLA-C", "HLA-E", "HLA-F", "HLA-G", "HLA-H")
HLA_CLASS_II_GENES <- c("HLA-DMA", "HLA-DMB", "HLA-DOA", "HLA-DOB", "HLA-DPA1", "HLA-DPB1", "HLA-DQA1", "HLA-DQA2", "HLA-DQB1", "HLA-DQB2", "HLA-DRA", "HLA-DRB1", "HLA-DRB3", "HLA-DRB4", "HLA-DRB5")

genes_class_i_present <- HLA_CLASS_I_GENES[HLA_CLASS_I_GENES %in% rownames(exp_mat_all_tpm_HLA)]
genes_class_ii_present <- HLA_CLASS_II_GENES[HLA_CLASS_II_GENES %in% rownames(exp_mat_all_tpm_HLA)]

MIN_CELLS_PER_SAMPLE_CELL_TYPE <- 5

# Filter to Pre-ICB and malignant cell types only (ALREADY excludes Malignant ND)
metadata_malignant_only <- 
  metadata_ICB %>%
  dplyr::filter(Rx == "Pre-ICB",
                Cell_type %in% c("AC-like", "MES-like", "GPC-like", "NPC-like", "OPC-like"))

# Step 1: Pseudobulk per SAMPLE per cell type (only if >5 cells)
bulk_expression_by_sample_cell_type <- 
  expand.grid(
    Sample_ID = unique(metadata_malignant_only$Sample_ID),
    Cell_type = c("AC-like", "MES-like", "GPC-like", "NPC-like", "OPC-like"),
    stringsAsFactors = FALSE
  ) %>%
  pmap_dfr(function(Sample_ID, Cell_type) {
    
    # Get all cell IDs for this sample and cell type
    cell_ids <- metadata_malignant_only %>%
      dplyr::filter(Sample_ID == !!Sample_ID, Cell_type == !!Cell_type) %>%
      pull(Cell_ID)
    
    # Filter: only include if >5 cells
    if (length(cell_ids) <= MIN_CELLS_PER_SAMPLE_CELL_TYPE) {
      return(NULL)
    }
    
    # Get patient ID for this sample
    patient_id <- metadata_malignant_only %>%
      dplyr::filter(Sample_ID == !!Sample_ID) %>%
      pull(Patient_ID) %>%
      unique()
    
    cat("Processing:", Sample_ID, "-", Cell_type, "(", length(cell_ids), "cells )\n")
    
    # Calculate mean expression across all cells of this sample-cell type
    exp_subset <- exp_mat_all_tpm_HLA[, cell_ids, drop = FALSE]
    mean_exp <- log2(Matrix::rowMeans(exp_subset) + 1)
    
    data.frame(
      gene_name = names(mean_exp),
      mean_exp = mean_exp,
      Sample_ID = Sample_ID,
      Patient_ID = patient_id,
      Cell_type = Cell_type,
      n_cells = length(cell_ids)
    )
  })

cat("\n=== Sample-cell type combinations after filtering (>5 cells) ===\n")
sample_cell_type_summary <- bulk_expression_by_sample_cell_type %>%
  dplyr::select(Sample_ID, Patient_ID, Cell_type, n_cells) %>%
  unique() %>%
  arrange(Cell_type, Patient_ID, Sample_ID)
print(sample_cell_type_summary)

# Step 2: Average sample-level pseudobulks to patient level
bulk_expression_by_patient_cell_type <- 
  bulk_expression_by_sample_cell_type %>%
  group_by(gene_name, Patient_ID, Cell_type) %>%
  summarise(
    mean_exp = mean(mean_exp, na.rm = TRUE),
    n_samples = n(),
    total_cells = sum(n_cells),
    .groups = "drop"
  )

cat("\n=== Patient-cell type combinations (after averaging samples) ===\n")
patient_cell_type_summary <- bulk_expression_by_patient_cell_type %>%
  dplyr::select(Patient_ID, Cell_type, n_samples, total_cells) %>%
  unique() %>%
  arrange(Cell_type, Patient_ID)
print(patient_cell_type_summary)

# Step 3: Average across patients for each cell type
bulk_expression_by_cell_type <- 
  bulk_expression_by_patient_cell_type %>%
  group_by(gene_name, Cell_type) %>%
  summarise(
    mean_exp = mean(mean_exp, na.rm = TRUE),
    n_patients = n_distinct(Patient_ID),
    total_cells = sum(total_cells),
    .groups = "drop"
  )

# Print summary
cat("\n=== Summary per cell type ===\n")
bulk_expression_by_cell_type %>%
  dplyr::select(Cell_type, n_patients, total_cells) %>%
  unique() %>%
  arrange(desc(n_patients)) %>%
  print()

# Create matrix for row-centering
bulk_expression_matrix <- 
  bulk_expression_by_cell_type %>%
  dplyr::select(gene_name, Cell_type, mean_exp) %>%
  pivot_wider(id_cols = gene_name, 
              names_from = Cell_type, 
              values_from = mean_exp) %>%
  column_to_rownames(var = "gene_name") %>%
  as.matrix()

# Row-center across all malignant cell types - USING scalop::rowcenter()
bulk_expression_centered_matrix <- scalop::rowcenter(bulk_expression_matrix)  # CHANGED THIS LINE

# Filter to HLA Class I and Class II genes
hla_class_i_matrix <- bulk_expression_centered_matrix[genes_class_i_present, , drop = FALSE]
hla_class_ii_matrix <- bulk_expression_centered_matrix[genes_class_ii_present, , drop = FALSE]

# Calculate average expression for Class I genes
class_i_average <- colMeans(hla_class_i_matrix, na.rm = TRUE)
class_i_average_row <- matrix(class_i_average, nrow = 1, dimnames = list("Average", names(class_i_average)))

# Calculate average expression for Class II genes
class_ii_average <- colMeans(hla_class_ii_matrix, na.rm = TRUE)
class_ii_average_row <- matrix(class_ii_average, nrow = 1, dimnames = list("Average", names(class_ii_average)))

# MODIFIED: Reorder HLA Class I genes to put B2M between HLA-G and Average
hla_class_i_genes_reordered <- setdiff(rownames(hla_class_i_matrix), "B2M")
b2m_row <- hla_class_i_matrix["B2M", , drop = FALSE]
hla_class_i_matrix_with_avg <- rbind(
  hla_class_i_matrix[hla_class_i_genes_reordered, , drop = FALSE],
  b2m_row,
  class_i_average_row
)

# Add average row to Class II matrix
hla_class_ii_matrix_with_avg <- rbind(hla_class_ii_matrix, class_ii_average_row)

# Remove duplicate B2M from Class II (only for display)
b2m_row_in_class_ii <- which(rownames(hla_class_ii_matrix_with_avg) == "B2M")
if (length(b2m_row_in_class_ii) > 0) {
  hla_class_ii_matrix_with_avg <- hla_class_ii_matrix_with_avg[-b2m_row_in_class_ii, , drop = FALSE]
}

# Combine Class I and Class II
hla_combined_matrix <- rbind(hla_class_i_matrix_with_avg, hla_class_ii_matrix_with_avg)

# MODIFIED: Define new cell type order
cell_type_order <- c("MES-like", "AC-like", "OPC-like", "NPC-like", "GPC-like")

# Reorder columns
hla_combined_matrix <- hla_combined_matrix[, cell_type_order, drop = FALSE]

# Define colors
malignant_colors <- c(
  "NPC-like" = "#003399",
  "OPC-like" = "#6699CC",
  "AC-like" = "#FFB366",
  "GPC-like" = "#FF8899",
  "MES-like" = "#CC0000"
)

# Create column annotation
col_ha <- HeatmapAnnotation(
  Cell_type = colnames(hla_combined_matrix),
  col = list(Cell_type = malignant_colors),
  annotation_name_side = "left",
  show_legend = FALSE,
  show_annotation_name = FALSE,
  border = TRUE
)

# Create row annotation for HLA class
row_annotation_df <- data.frame(
  gene_name = rownames(hla_combined_matrix)
) %>%
  mutate(HLA_class = case_when(
    gene_name == "Average" & row_number() <= (nrow(hla_class_i_matrix_with_avg)) ~ "HLA Class I",
    gene_name == "Average" ~ "HLA Class II",
    gene_name %in% genes_class_i_present ~ "HLA Class I",
    gene_name %in% genes_class_ii_present ~ "HLA Class II"
  ))

# Create color function for heatmap
col_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#2166AC", "#92C5DE", "#F7F7F7", "#D6604D", "#B2182B")
)

# Create heatmap
Heatmap(
  hla_combined_matrix,
  name = "Relative\nExpression",
  col = col_fun,
  
  # Column settings
  cluster_columns = FALSE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 10, fontface = "italic"),
  column_names_rot = 45,
  column_names_side = "bottom",
  top_annotation = col_ha,
  column_title = NULL,
  border = TRUE,
  
  # Row settings
  cluster_rows = FALSE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 9, fontface = "italic"),
  row_names_side = "right",
  row_split = row_annotation_df$HLA_class,
  row_title_gp = gpar(fontsize = 11, fontface = "bold"),
  row_title_rot = 0,
  row_gap = unit(2, "mm"),
  
  # Cell aesthetics
  rect_gp = gpar(col = "white", lwd = 1),
  
  # Heatmap body
  heatmap_legend_param = list(
    title = "Relative\nExpression",
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    at = c(-2, -1, 0, 1, 2),
    labels = c("-2", "-1", "0", "1", "2"),
    labels_gp = gpar(fontsize = 9),
    legend_height = unit(3, "cm"),
    legend_width = unit(0.4, "cm"),
    border = "black", 
    width = unit(6, "cm"),    # Decrease this for thinner (was 8)
    height = unit(16, "cm")   # Increase this for taller (was 12)
    
  ),
  
  # Size
  width = unit(8, "cm"),
  height = unit(18, "cm")
)

```
